---
title: "Live session 2.4: Residual diagnostics"
author: "Matthew Ragoza"
date: "1/28/2022"
always_allow_html: true
output:
  html_document:
    df_print: paged
    toc: false
    number_sections: false
    toc_depth: '3'
    code_folding: show
---

## Read in the data.

```{r}
library('dplyr')
library('tidyverse')
library('reshape2')
library('ggplot2')
library('gridExtra')

setwd('C:\\Users\\mtr22\\Code\\BIOST2049\\module-2\\live-sessions')
data <- read.csv('Module1Topics1and2Example.csv')
data
```
## Descriptive statistics

```{r}
desc <- data %>%
  melt(id.vars='patient') %>%
  group_by(variable) %>%
  summarize(
    mean=mean(value),
    SD=sd(value),
    Q1=quantile(value, 0.25),
    Q2=quantile(value, 0.50),
    Q3=quantile(value, 0.75),
    IQR=IQR(value),
    min=min(value),
    max=max(value)
  ) %>%
  mutate_if(is.numeric, round, digits=1)

y_desc  <- filter(desc, variable=='y')
x1_desc <- filter(desc, variable=='x1')
x2_desc <- filter(desc, variable=='x2')
x3_desc <- filter(desc, variable=='x3')
desc
```

## Data visualization

```{r, fig.width=10.5, fig.height=4.5, message=FALSE, warning=FALSE}

blank <- ggplot() + theme_void()

# axes scales
y_limits  = c(4, 72)
x1_limits = c(2, 4)
x2_limits = c(5, 9)
x3_limits = c(0, 32)

# axes ticks
y_ticks  = seq(y_limits[1],  y_limits[2], 10)
x1_ticks = seq(x1_limits[1], x1_limits[2], 0.5)
x2_ticks = seq(x2_limits[1], x2_limits[2], 1)
x3_ticks = seq(x3_limits[1], x3_limits[2], 5)

# bin widths
y_binwidth = (y_limits[2] - y_limits[1])/20
x1_binwidth = (x1_limits[2] - x1_limits[1])/20
x2_binwidth = (x2_limits[2] - x2_limits[1])/20
x3_binwidth = (x3_limits[2] - x3_limits[1])/20

# histograms

plot_y <- data %>%
  ggplot(aes(y=y)) +
  geom_histogram(binwidth=y_binwidth) +
  theme_classic() +
  theme(panel.grid.major=element_line()) +
  scale_y_continuous(breaks=y_ticks, limits=y_limits)

plot_x1 <- data %>%
  ggplot(aes(x=x1)) +
  geom_histogram(binwidth=x1_binwidth) +
  theme_classic() +
  theme(panel.grid.major=element_line()) +
  scale_x_continuous(breaks=x1_ticks, limits=x1_limits)

plot_x2 <- data %>%
  ggplot(aes(x=x2)) +
  geom_histogram(binwidth=x2_binwidth) +
  theme_classic() +
  theme(panel.grid.major=element_line()) +
  scale_x_continuous(breaks=x2_ticks, limits=x2_limits)

plot_x3 <- data %>%
  ggplot(aes(x=x3)) +
  geom_histogram(binwidth=x3_binwidth) +
  theme_classic() +
  theme(panel.grid.major=element_line()) +
  scale_x_continuous(breaks=x3_ticks, limits=x3_limits)

# scatter plots

plot_yx1 <- data %>%
  ggplot(aes(x=x1, y=y)) +
  geom_point() +
  geom_smooth(method='lm', color='black') +
  theme_classic() +
  theme(panel.grid.major=element_line()) +
  scale_y_continuous(breaks=y_ticks, limits=y_limits) +
  scale_x_continuous(breaks=x1_ticks, limits=x1_limits)

plot_yx2 <- data %>%
  ggplot(aes(x=x2, y=y)) +
  geom_point() +
  geom_smooth(method='lm', color='black') +
  theme_classic() +
  theme(panel.grid.major=element_line()) +
  scale_y_continuous(breaks=y_ticks, limits=y_limits) +
  scale_x_continuous(breaks=x2_ticks, limits=x2_limits)

plot_yx3 <- data %>%
  ggplot(aes(x=x3, y=y)) +
  geom_point() +
  geom_smooth(method='lm', color='black') +
  theme_classic() +
  theme(panel.grid.major=element_line()) +
  scale_y_continuous(breaks=y_ticks, limits=y_limits) +
  scale_x_continuous(breaks=x3_ticks, limits=x3_limits)

# arrange plots on grid

grid.arrange(
  plot_x1,  plot_x2,  plot_x3,  blank,
  plot_yx1, plot_yx2, plot_yx3, plot_y,
  nrow=2, heights=c(1.5, 3),
  ncol=4, widths=c(3, 3, 3, 1.5)
)
```


## Fit linear regression model with interaction term.

```{r}
model <- lm(sbp ~ age + sex + age*sex, data=data)
```

## Perform partial F test on interaction coefficient.

```{r}
summary(model)
anova(model)
```
The p-value of 0.9342 is not less than $\alpha$=0.05, so we do not reject the null hypothesis. We do not have evidence of a significant interaction.

## Create an interaction plot to check parallelism.

```{r, fig.width=5, fig.height=4}
ggplot(
  data=data,
  aes(x=age, y=sbp, group=sex)
) +
  geom_point(aes(shape=sex)) +
  geom_smooth(method='lm', aes(linetype=sex), color='black') +
  theme_classic()
```

The interaction plot supports the conclusion of the hypothesis test, as it appears that the lines are parallel for the different sex categories.

## Perform partial F test for coincidence.

```{r}
reduced_model = lm(sbp ~ age, data=data)
anova(reduced_model, model)
```

## Check for confounding

```{r}
model = lm(sbp ~ sex, data=data)
adjusted_model = lm(sbp ~ sex + age, data=data)
summary(model)
summary(adjusted_model) # check whether sex coefficient changed >= 10%
```

TODO run margins command to get unadusted sbp means by sex
