---
title: "Live session 1.3: Confounding and interactions"
author: "Matthew Ragoza"
date: "1/21/2022"
always_allow_html: true
output:
  html_document:
    df_print: paged
    toc: false
    number_sections: false
    toc_depth: '3'
    code_folding: show
---

## Read in the data.

```{r}
library('ggplot2')
setwd('C:\\Users\\mtr22\\Code\\BIOST2049\\module-1\\live-sessions')
data <- read.csv('Module1Topic3Example.csv')
data$sex <- factor(data$sex, labels=c('male', 'female'))
data
```

## Fit linear regression model with interaction term.

```{r}
model <- lm(sbp ~ age + sex + age*sex, data=data)
```

## Perform partial F test on interaction coefficient.

```{r}
summary(model)
anova(model)
```
The p-value of 0.9342 is not less than $\alpha$=0.05, so we do not reject the null hypothesis. We do not have evidence of a significant interaction.

## Create an interaction plot to check parallelism.

```{r, fig.width=5, fig.height=4}
ggplot(
  data=data,
  aes(x=age, y=sbp, group=sex)
) +
  geom_point(aes(shape=sex)) +
  geom_smooth(method='lm', aes(linetype=sex), color='black') +
  theme_classic()
```

The interaction plot supports the conclusion of the hypothesis test, as it appears that the lines are parallel for the different sex categories.

## Perform partial F test for coincidence.

```{r}
reduced_model = lm(sbp ~ age, data=data)
anova(reduced_model, model)
```

## Check for confounding

```{r}
model = lm(sbp ~ sex, data=data)
adjusted_model = lm(sbp ~ sex + age, data=data)
summary(model)
summary(adjusted_model) # check whether sex coefficient changed >= 10%
```

TODO run margins command to get unadusted sbp means by sex
